### 2018-04-04 concepts.tex
bib

@@

In \emph{monotonic} cases, removing a part of a test cannot
increase the probability of a predicate holding.  This is fairly
common.  In these cases, when use a reduction algorithm to reduce $t$
to $r$, $P(p(r) \leq P(p(t))$.  In \emph{non-monotonic} cases,
however, there is no such upper bound.  Removing a step in $R$

@@

>>>>

\subsubsection{Determinism and Reflexive Differential Testing}

Horizontal determinism can be best understood by thinking of
nondeterminism detection as an unusual kind of \emph{differential
  testing} \cite{Differential,ICSEDiff}.  In differential testing, a
system is compared against a reference in order to ensure that
it behaves equivalently, at some level of abstraction, to another
implementation of the same functionality.  Differential testing is
extremely powerful, in that any divergence of behavior, if divergence
is correctly defined, indicates a functional correctness fault in (at
least) one of the systems under test.  Being able to detect functional
correctness errors without the cost of constructing a formal
specification is extremely useful in automated test generation.  Differential testing is widely
used for systems software components such as compilers
\cite{Differential,csmith} and POSIX file systems \cite{CFV08,AMAI}, where multiple
implementations are common.  The major limitation of differential
testing, of course, is that multiple implementations of a system are
almost as rare as good correcteness specifications.

For the special case of detecting nondeterminism, however, \emph{a
  system can serve as its own reference implementation.}

@@

Delta-debugging \cite{DD}\footnote{We use delta-debugging to stand for
  all test reduction algorithms, even those \cite{CReduce,tstl} that do not use the exact
  binary-search that distinguishes delta-debugging \emph{per se}; the
  differences are immaterial for our purposes.}  is a widely used method for reducing the
size of failing tests, making them easier to understand and debug.
Delta-debugging in the context of detecting nondeterminism has two
purposes.  One is simply the usual goal of reducing the size of a
test.  Identifying the cause of nondeterminism may be very easy in a
test consisting of ten library function calls (it is one of these ten
calls), but very difficult in a test consisting of a hundred library
function calls.  This is no different than the common use of
delta-debugging.  However, in nondeterminism detection,
delta-debugging can also be used to \emph{change the probability of
  nondeterministic behavior}.

In \emph{monotonic} cases, removing a part of a test cannot
increase the probability of a predicate holding.  This is fairly
common.  In these cases, when we use a reduction algorithm to reduce $t$
to $r$, $P(p(r) \leq P(p(t))$.  In \emph{non-monotonic} cases,
however, there is no such upper bound.  Removing a step in $t$ may
increase the probability that $t$ behaves nondeterministically.

@@

The only absolutely necessary changes required to use standard
delta-debugging implementations in nondeterminism detection
are simply removal of some ``sanity checks'' in the code:  many
implementations (including the Python code provided by Zeller) assert
that the predicate holds on the original test.  In probabilistic
settings, this may not be true, and we may even be trying to find a
subset where a predicate that is very far from holding on the original test
holds (in a non-monotonic case where we aim to increase the probability
of nondeterminism).

\subsubsection{``Publication Bias'' in Delta-Debugging}

However, simply using delta-debugging off the shelf with a predicate
like $P(fail) > 0.3 \wedge P(fail) < 1.0$, to force a test to be flaky, and force it to fail
sufficiently often to be used in debugging, will often produce
surprising and unfortunate results.  In many cases,
delta-debugging will indeed reduce the large test to a small subset.
And, in a technical sense, delta-debugging will work:  it will never
convert a nondeterministic test to a completely deterministic test,
because the reduced test that delta-debugging returns is always one
where the predicate of interest has been seen to evaluate to true.
However, if you run the resulting test, it will, in many cases, have a
$P(fail)$ that is much, much smaller than
0.3, perhaps as low as 0.01.  Why?

The problem is analogous to the problem of publication bias in
scientific fields \cite{ahmed2012assessment}.  A predicate like
$P(fail) > 0.3 \wedge P(fail) < 1.0$ cannot be evaluated by
determining the true probability of failure; rather, the test must be
run some concrete number of times, and the number of failures counted.
However, even if the number of samples $N$ is large, there is some
probability (based on the sample size) of a result that diverges
significantly from the actual probability of failure.  If the
predicate were run only once, and the number of samples reasonably
large, this would not matter.  However, by their nature,
delta-debugging and other test reduction algorithms, explore a search
space that often contains hundreds or even thousands or millions of
candidate tests.  The predicate is evaluated on each of these, and so
even with large $N$, it is extremely likely that some evaluation will
produce a very poor estimate of $P(fail)$.  If such an evaluation
causes the predicate to appear to hold for a test, delta-debugging is
``stuck'' with the error, because (to our knowledge) no test reduction
algorithms allow backtracking.  After such a mistake, finding further
reductions will become harder, but may still be possible due to the
same source of errors:  the number of experiments run is far larger
than the probability of an incorrect result.

It is the combination of
a one-way bias on faulty evaluations (the consequences of a false
positive for the predicate are much greater than for a false negative)
and the huge number of experiments relative to error rate, that
produces bad results, akin to the magnification of effect sizes
in science due to publication bias.  The bias in delta-debugging tends
towards producing a reduced test where probabilities are much smaller
than demanded by a predicate, due to the pressure on delta-debugging
to produce shorter tests.  Shorter tests have smaller probabilities,
on average, due to two factors:  first, timing-induced nondeterminism
has a much smaller temporal space to operate in, and second, if some
test operation introduces a small probability of nondeterminism, and
only many repetitions of that operation make the probability large,
small tests obviously on average have fewer instances of the
problematic operation.

In scientific literature, the most frequently proposed solution is the
use of replications:  repeated runs of ``successful'' experiments to
minimize the probability that a result is a fluke due to publication
bias.  One way to produce this effect would be to allow
delta-debugging to backtrack if the probabilties observed in predicate
evaluations suddely exhibit a strong discontinuity, a kind of ``paper
retraction'' based on near-replications.  However, this requires
modifying the delta-debugging implementations, which is difficult and
sometimes not really feasible (e.g., few users of AFL-fuzz \cite{aflfuzz} will wish
to change its C code for test reduction).  Ideally, the solution
should be implementable simply by modifying the predictate that is
evaluated.

A costly but effective solution is to make $N$ large in comparison to
the number of expected predicate evaluations performed during
delta-debugging.  If the error rate is low enough, the problem
disappears.  However, given the large number of evaluations performed,
this will tend to make delta-debugging extremely slow.  We propose
using a dynamic sampling approach, where $N$ is small, but if the
predicate evaluates to true, $M$ repeated true evaluations
(``replications'') are
required before the predicate returns ``true'' to the delta-debugging
algorithm.  Tuning $M$, any degree of confidence can be achieved, with
the basic tradeoff being between finding a test with the desired
probabilities, and the speed and effectiveness of delta-debugging.
Because increasing $M$ makes false negatives more likely, larger $M$
will usually result in less-than-optimal reduction of the original test.

### 2018-04-04 implementation.tex
3.3 hashing
edits
reorg
bib

>>>>

Because TSTL supports differential testing \cite{tstlsttt}, horizontal
nondeterminism detection can technically be implemented simply by
declaring a system to be its own reference, using TSTL's notation for
differential testing.  However, such an approach requires some effort
on the part of the user to control which values are checked for
equivalence, and does not easily support injecting timing differences,
or using a new process, in checking the behavior.

### 2018-04-04 introduction.tex
bib

WHY DETERMINISM:  Bugs, flaky tests, pain debugging.

@@

A system is deterministic if, given its complete state at a point in
time, it is possible, in principle, to predict its future behavior
perfectly.  We say ``in principle'' because in the real world,
prediction may be possible but hopelessly impractical.  We write
complex software systems in many cases because we cannot predict their
behavior (if we could perform the calculations in advance, ourselves,
we would just do so).  As a consequence, in software, rather than
defining determinism in terms of prediction, we usually therefore
simply say that a system is deterministic if, given the same state and
inputs, it always produces the same outputs\footnote{Of course, this
  shift of definition is often adopted in arguments for physical or
  human behavioral determinism, due to the even greater difficulties
  of prediction in those arenas.}.  The change of
definition, we will see, is central to our approach, and comes with a
number of limitations.

@@

Unexpected nondeterminism usually arises when there is an element of
state or input that \emph{is} critical to the produced behavior, but
the programmer has not anticipated.  E.g., it is beleived that the
behavior of a thread scheduler will not matter, but a race condition
in the code means that it does matter, after all.

@@

Unexpected nondeterminism is, unfortunately, usually only discovered
in a context that makes it very hard to debug.   The most common such
contexts are occasional rare failures of a system in deployment, and
flaky tests.

>>>>

Determinism is particularly important for testing and debugging, where
being able to exactly reproduce system behavior is essential to
productivity.  If executing ``the same test'' can produce
significantly different behavior each time it is run, the consequences
can be unfortunate.  Developers using a test exhibiting nondeterminism
to debug a system
face a challenge in reasoning about causality, in that an event
observed may not even take place the next time the test is run.
Automated test generation systems using test coverage results to drive
the search for interesting inputs may be mislead when a run executes
code only rarely.  And, most significantly, regression testing
effectiveness can be significantly reduced if tests fail only
intermittently and unpredictably, as a result of environmental
factors rather than bugs in changed code.  Such behavior is
unfortunately all too common:  Gao et
al. \cite{Gao:2015:MSU:2818754.2818764} observed coverage differences
of up to 184 lines of code for the same test, and false positive rates
as high as 96\%.  Nondeterminism is sometimes problematic for
developers, who will usually want to  assume that library code they
use is deterministic; it is often disastrous for large-scale or highly
automated testing.

@@

A system is deterministic if, given the system's complete state at a point in
time, it is possible, in principle, to predict its future behavior
perfectly.  We say ``in principle'' because in the real world,
prediction may be possible but hopelessly impractical.  We write
complex software systems in many cases because we cannot predict their
behavior (if we could perform the calculations in advance, ourselves,
we would just do so).  As a consequence, in software, rather than
defining determinism in terms of prediction, we usually therefore
simply say that a system is deterministic if, given the same state and
inputs, it always produces the same outputs\footnote{Of course, this
  shift of definition is often adopted in arguments for physical or
  human behavioral determinism, due to the even greater difficulties
  of prediction in those arenas.}.  The change of
definition, we will see, is central to our approach, and comes with a
number of limitations.

@@

Because the
programmer's abstraction of the system ignores  such a large number of
details, very few tests are in fact ``deterministic'' in the sense
that they eliminate all changes in behavior between executions.  Running the same test
twice almost always results in differences, given a low enough level of
abstraction, since the system load,
cache contents, branch predictor history, etc. are almost never
controlled for; however, this kind of
nondeterminism is usually of no interest, unless the test involves
extremely precise timing constraints.  Rather, what matters is when
some kind of nondeterminism unexpectedly impacts computed values at a
higher level of abstraction; in general, if the operating system
itself is not buggy, low-level nondeterminism, by design, is invisible
except in fine-grained performance testing or real-time systems.
Unexpected nondeterminism usually arises when there is an element of higher-level
state or input that \emph{is} critical to the produced behavior, but
the programmer has not anticipated.  E.g.,  when it is believed that the
behavior of a thread scheduler will not matter, but a race condition
in the code means that it does matter, or when the order of items in
an iterator on a hash table is important, and the hash values used are
salted and thus not predictable.

@@

Unexpected nondeterminism is, unfortunately, usually only discovered
in a context that makes it very hard to debug.   The most common such
contexts are occasional rare failures of a system in deployment, and
regression tests that do not behave reliably (known as flaky tests).
Furthermore, unexpected nondeterminism makes it difficult to use
automated test generation to produce effective regression tests for a
system.  While developers may know how to produce reliably
deterministic unit tests, even in the presence of underlying
nondeterminism, automated test generation, without a large investment
in human time, usually does not have sufficient information to avoid
producing the occasional such test.  Moreover, where a human might
make an assertion about the final value produced during a unit test,
an automatically generated regression test is usually most easily produced by
simply asserting equality with all values produced during a reference
run, since the final step of a test
is likely to be somewhat arbitrary, generated in an effort to increase
code coverage, not establish some functional correctness property.
This may be one reason that automated testing is seldom used to
produce general regression tests.  While failure-inducing tests
produced automatically are often added to regression suites, the full
set of passing tests is, to our knowledge, only infrequently preserved
as part of a typical regression suite, perhaps because such tests are
likely to be \emph{flaky} without considerable human effort.

\subsubsection{Nondeterminism and Flaky Tests}

In order to help ensure that they are reliable and secure, complex
modern software systems usually include a large set of
\emph{regression tests}.  A regression test suite is a (usually large)
set of tests that can be run against a software system every time it
is modified, to ensure that the modification has not broken the system
in some way.  \emph{Flaky tests} \cite{miccoflaky} are regression
tests that fail in an intermittent, unreliable fashion, and thus
degrade the utility of regression testing.  The essence of a flaky
test is that, for the same snapshot of test code and code under test
(CUT), it sometimes fails and sometimes passes: the pass/fail result (disposition) of the
test is not a deterministic property of the test code, code under
test, and environment running the test.  This produces three serious
problems: first, a flaky test often wastes developer time and delays
software changes by forcing the investigation of code-under-test that
is not actually incorrect.  Second, failures in flaky tests (for that
reason) are often ignored, and therefore serious software faults may
be missed.  Finally, to mitigate these problems, flaky tests are often
run multiple times, wasting valuable computing resources and also
delaying acceptance of code changes.  A canary in a coal mine is of
little use if canaries frequently become ill for reasons unrelated to
the presence of toxic gases.  Mining may stop for no good reason, or
miners may learn to ignore the canary, leading to tragedy; a third,
more ``practical'' option is that miners may carry so many redundant
canaries into the coal mine that canary-care becomes a serious burden
on mining.

Flaky tests are, for us, simply a special case of general test
nondeterminism, where the nondeterminism of test values is sufficient
to cause the pass/fail result of the test to vary.  The definitions
and techniques proposed in this paper all apply to flaky tests, as a
special case of various kinds of nondeterminism described below.

\subsection{Contributions:  Detecting and Debugging Nondeterminism}

This paper proposes (1) a number of formal definitions of types of
nondeterminism and (2) a number of techniques, based on these definitions, for detecting and debugging
nondeterminism in property-based testing, particularly for library
code.  The methods also work for testing at the system level, but are
perhaps most effective for identifying the low-level sources of
nondeterminism in libraries.  This enables a number of useful results:
 the automated production of
nondeterminism-free regression tests even for libraries with some
sources of nondeterminism, the documentation of library
nondeterminism, and, most importantly, detection and effective debugging of
unexpected nondeterminism that is, itself, a bug.  We first
describe the types of nondeterminism we aim to detect, and the
extension of a widely used debugging tool to better handle
nondeterminism as a test property, then describe an implementation of
these ideas in a Python testing tool that is used on real-world
projects.  Finally, we use a set
of case studies of real-world Python libraries to demonstrate the
utility of our ideas and understand the cost of adding nondeterminism
detection to existing automated testing.

### 2018-04-04 relatedwork.tex
bib

Shi et. al \cite{DetermImp} examined what might be seen as a related, though in a sense the opposite, problem:  they detect code that assumes a deterministic implementation of a non-deterministic specification.  E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order produced by a given implementation.

>>>>

Gao et al. generally considered the question of how to make tests repeatable \cite{Gao:2015:MSU:2818754.2818764}, in the context of system-wide user interaction tests.  Their work focused on systemic factors, such as execution platform and persistent configuration and database files, in contrast to our focus on identifying nondeterminism at the library level.  However, what they refer to as ``test harness factors'' includes delays, which can be of importance to nondeterminism at the library level when asynchronous behavior is involved.

Shi et. al \cite{DetermImp} examined what might be seen as a related, though in a sense the opposite, problem:  they detect code that assumes a deterministic implementation of a non-deterministic specification.  E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order produced by a given implementation.  This procedure, applied to Python code in a pre-3.3 environment, would have flagged many instances of the nondeterminism that arose on the introduction of salted hashing.  Determinism is also sometimes used as a property in domain-specific testing efforts, e.g. in testing shader compilers \cite{shader}.

The problem of test nondeterminism is closely related to the (extremely important in industrial settings) issue of flaky tests \cite{miccoflaky, luo2014empirical,palomba2017does,listfieldtestanalysis}.  How to handle flaky tests in practice (when they cannot be avoided) is a major issue in Google-scale continuous testing \cite{memon2017taming}, and, as Memon et al. describe, the problem of flaky tests influences the general design of Google test automation.
Previous work on flaky tests has either focused on test inter-dependence as a cause of flaky behavior \cite{LamZE2015}, or provided large-scale empirical examination of tests from one large open source project (Apache) \cite{luo2014empirical,palomba2017does}.  Palomba and Zaidman \cite{palomba2017does} investigated the relationship between test code smells and flaky tests.   The present paper, rather than focusing on flaky tests per se, investigates new tools for handling nondeterminism in property-based test generation.  From our point of view, flaky behavior is simply a special case of nondeterminism, where the nondeterminism is sufficient to cause the test to have different pass/fail results at times.
